### 2019-04-23 initial/notes.tex
maybe

This Hamiltonian is Kassandras Eq.\ 13, in the co-rotating frame with the
perturber's angular momentum:
\begin{equation}
    \mathcal{H} = \frac{1}{2}\p*{\hat{s} \cdot \hat{l}}^2
        - \eta \p*{\hat{s} \cdot \hat{l}_p}.
\end{equation}
In this frame, we can choose $\hat{l} \equiv \hat{z}$ fixed, and $\hat{l}_p =
\cos I\hat{z} + \sin I\hat{x}$ fixed as well. Then
\begin{equation*}
    \hat{s} = \cos\theta \hat{z}
        - \sin\theta\p*{\sin \phi \hat{y} + \cos \phi \hat{x}}.
\end{equation*}
We can choose the convention for $\phi = \phi$ azimuthal angle requiring $\phi =
0, \pi$ mean coplanarity between $\hat{s}, \hat{l}, \hat{l}_p$ in the $\hat{x},
\hat{z}$ plane such that $\hat{l}_p, \hat{s}$ lie on the same side of
$\hat{l}$. Then we can evaluate in coordinates
\begin{align*}
    \hat{s} \cdot \hat{l} &= \cos \theta,\\
    \hat{s} \cdot \hat{l}_p
        &= \cos \theta \cos I - \sin I \sin \theta \cos \phi,\\
    \mathcal{H} &= \frac{1}{2}\cos^2\theta
        - \eta \p*{\cos \theta \cos I - \sin I \sin \theta \cos \phi}.
\end{align*}
Note that if we take $\cos\theta$ to be our canonical variable, $\sin\theta =
\sqrt{1 - \cos^2\theta}$ can be used.

@@

Alternatively, consider Hamilton's equations applied to the Hamiltonian:
\begin{align}
    \pd{\phi}{t} = \pd{\mathcal{H}}{(\cos\theta)}
        &= \cos\theta - \eta\p*{\cos I + \sin I \cot \theta \cos \phi},\\
    \pd{(\cos \theta)}{t} = -\pd{\mathcal{H}}{\phi}
        &= +\eta \sin I \sin \theta \sin \phi.
\end{align}
This produces the same trajectories as the Cartesian EOM, so this is correct.
However, since $\pd{\phi}{t} \propto 1/\sin\theta$, this is not a desirable
system of equations to use, as they are very stiff near $\theta \approx 0$.

@@

We notate $\mu = \cos\theta$; note that CS4 is $\mu_4 \approx \frac{\eta \cos
I}{1 + \eta \sin I} \approx \eta \cos I$. Setting the Hamiltonian equal to its
value at CS4 gives
\begin{align*}
    H_4 &\equiv H\p*{\mu_4, \phi_4}
        \approx \frac{\mu_4^2}{2} - \eta \mu_4 \cos I + \eta \sin I,\\
        &= -\eta^2\cos^2 I + \eta \sin I,\\
    H(\mu_{sep}, \phi_{sep})
        &= H_4 = \eta \sin I \cos \phi + \frac{\mu^2}{2} - \eta \mu \cos I
            + \mathcal{O}(\eta^3),\\
    0 &\approx \frac{\mu_{sep}^2}{2} - \eta \mu_{sep} \cos I
        - \eta \sin I\p*{1 - \cos \phi_{sep}} + \eta^2\cos^2I,\\
    \mu_{sep}\p*{\phi} &\approx \sqrt{2\eta \sin I\p*{1 - \cos \phi}}
        + \mathcal{O}(\eta).
\end{align*}

@@

We zoom in on Cassini State $4$, which has $\cos \theta \approx 0, \phi = 0$. In
particular, by using equations of motion
\begin{align}
    \pd{\phi}{t} &= \cos \theta - \eta\p*{\cos I + \sin I \cot \theta \cos \phi}
        ,\\
    \pd{\theta}{t} &= -\eta \sin I \sin \phi - \s*{\epsilon \sin \theta},
\end{align}
we can compute $\theta_4 \approx -\frac{\pi}{2} +\frac{\eta \cos I}{1 + \eta
\sin I}, \phi_4 = 0$ in the $\epsilon = 0$ limit (the other solution is
$\theta_2 \approx +\frac{\pi}{2} - (\dots)$). Indeed the signs are correct:
near $\theta \approx \frac{\pi}{2}, \phi \approx 0$ we have EOM
$\dot{\delta\phi}_4 \approx -\delta \theta_4$ while $\dot{\delta \theta_4}
\approx -\eta \sin I \delta \phi_4$ which is unstable.

Substituting in perturbatively for nonzero $\epsilon$ in $\pd{\theta}{t}$ gives
$\delta \phi_4 = +\frac{\epsilon}{\eta \sin I}, \delta \theta_4 = 0$. This is in
agreement with Dong's result. Note that $\delta \theta_2 = -\frac{\epsilon}{\eta
\sin I}$, which I saw in my simulations.

@@

\subsection{Melnikov Distance}

We notice that the separatrix is a heteroclinic orbit, or a saddle connection,
in the dissipation free problem. Introducing dissipation breaks the saddle
connection by a distance that can be estimated with the Melnikov distance. This
is G\&H Equation 4.5.11 or something:
\begin{align}
    d(t_0) &= \frac{\epsilon M(t_0)}{\abs*{f(q^0(0))}} + \mathcal{O}(\epsilon^2)
        ,\\
    M(t_0) &= \int\limits_{-\infty}^\infty
        \s*{f \times g}_{hetero}\;\mathrm{d}t.
\end{align}
We evaluate the Melnikov integral $M(t_0)$ on the heteroclinic orbit. Note that
since in our problem our perturbation $g$ is time-independent, so too is the
Melnikov integral $M(t_0) = M$.

Let's apply this to the Cassini state Hamiltonian w/ dissipation. We first write
down our EOM in Melnikov form (we use canonical variables $\mu, \phi$):
\begin{align}
    \rd{\hat{s}}{t}
        &= \underbrace{
            \pd{\mathcal{H}}{\mu}\hat{\phi} - \pd{\mathcal{H}}{\phi}\hat{\mu}}_f
            + \epsilon \underbrace{\p*{1 - \mu^2}\hat{\mu}}_g.
\end{align}
Then $f \times g = f_\phi g_\mu = \pd{\mathcal{H}}{\mu}\p*{1 - \mu^2}$. We then
want to integrate this along the heteroclinic orbit. We can make change of
variables
\begin{equation}
    M = \int\limits_0^{2\pi}\pd{\mathcal{H}}{\mu}
        \p*{1 - \mu^2}\p*{\pd{\phi}{t}}^{-1}\;\mathrm{d}\phi.
\end{equation}
But thankfully, $\pd{\mathcal{H}}{\mu} = \pd{\phi}{t}$ in the absence of
dissipation, and so $M = 2\pi\p*{1 - \mu^2} \approx 2\pi\p*{1 - 2\eta \sin I}$.
Thus, the Melnikov distance at point $q^0$, a point on the heteroclinic
orbit of the unperturbed Hamiltonian, is just
\begin{equation}
    d(q^0) = \frac{2\pi \epsilon\p*{1 - 2\eta \sin I}}{\abs*{f(q^0)}}.
\end{equation}

We compute the maximum $\abs*{f(q^0)}$, which occurs at $\phi = \pi$, and we
find the expected $f \approx \sqrt{4\eta \sin I}$.

>>>>

This Hamiltonian is Kassandras Eq.\ 13, in the co-rotating frame with the
perturber's angular momentum:
\begin{equation}
    \mathcal{H} = \frac{1}{2}\p*{\hat{s} \cdot \hat{l}}^2
        - \eta \p*{\hat{s} \cdot \hat{l}_p}.
\end{equation}
In this frame, we can choose $\hat{l} \equiv \hat{z}$ fixed, and $\hat{l}_p =
\cos I\hat{z} + \sin I\hat{x}$ fixed as well. Then
\begin{equation*}
    \hat{s} = \cos\theta \hat{z}
        - \sin\theta\p*{\sin \phi \hat{y} + \cos \phi \hat{x}}.
\end{equation*}
We can choose the convention for $\phi = \phi$ azimuthal angle requiring $\phi =
0, \pi$ mean coplanarity between $\hat{s}, \hat{l}, \hat{l}_p$ in the $\hat{x},
\hat{z}$ plane such that $\hat{l}_p, \hat{s}$ lie on the same side of
$\hat{l}$. Then we can evaluate in coordinates
\begin{align*}
    \hat{s} \cdot \hat{l} &= \cos \theta,\\
    \hat{s} \cdot \hat{l}_p
        &= \cos \theta \cos I - \sin I \sin \theta \cos \phi,\\
    \mathcal{H} &= -\frac{1}{2}\cos^2\theta
        + \eta \p*{\cos \theta \cos I - \sin I \sin \theta \cos \phi}.
\end{align*}
Note that if we take $\cos\theta$ to be our canonical variable, $\sin\theta =
\sqrt{1 - \cos^2\theta}$ can be used.

@@

Alternatively, consider Hamilton's equations applied to the Hamiltonian:
\begin{align}
    \pd{\phi}{t} = \pd{\mathcal{H}}{(\cos\theta)}
        &= -\cos\theta + \eta\p*{\cos I + \sin I \cot \theta \cos \phi},\\
    \pd{(\cos \theta)}{t} = -\pd{\mathcal{H}}{\phi}
        &= -\eta \sin I \sin \theta \sin \phi.\label{eq:H_eom}
\end{align}
This produces the same trajectories as the Cartesian EOM, so this is correct.
However, since $\pd{\phi}{t} \propto 1/\sin\theta$, this is not a desirable
system of equations to use, as they are very stiff near $\theta \approx 0$.

\subsection{Cassini States}

The zeros to \autoref{eq:H_eom} are the Cassini states; we will go to canonical
variables $\mu = \cos\theta$. We can immediately see that $\sin\phi = 0$ is
necessary, so $\cos \phi = \pm 1$ and we need only solve for $\pd{\phi}{t} = 0$.
We can furthermore separate the problem into two regimes, $\eta \ll 1$ and $\eta
\gg 1$.

For $\eta \ll 1$, it is clear that there will be two solutions near $\mu^2 = 1$
and two solutions near $\mu = 0$:
\begin{itemize}
    \item For $\mu = 1 - \frac{\theta^2}{2}$, the dominant terms are
        $\pd{\phi}{t} \approx -1 + \eta \sin I \frac{1}{\theta} = 0$,
        where we've taken $\cos \phi = +1$ and $\phi = 0$. This forces $\theta
        = \eta \sin I$.

    \item Similarly, for $\mu = -1 + \frac{\epsilon^2}{2}$, $\phi = 0$ and
        $\epsilon = \eta \sin I$ again. This actually corresponds to $\theta =
        \pi - \eta \sin I$.

    \item For $\mu \approx 0$, we have instead $\pd{\phi}{t} = -\mu\p*{1 - \eta
        \sin I \cos \phi} + \eta \cos I = 0$. This forces $\mu_\pm = \frac{\eta
        \cos I}{1 \pm \eta \sin I}$, where $\phi_{\pm} = \pi, 0$ respectively.

        Note that $\phi = 0, \mu \approx 0$ is conventionally CS4. The
        linearization locally has form $\pd{\delta \phi}{t} = -\delta \mu\p*{1 -
        \eta \sin I}$ and $\pd{\mu}{t} = -\eta \sin I$, so the eigenvalues are
        $\approx \mp \sqrt{\eta \sin I}$, and the two eigenvectors are $\p*{1,
        \pm \sqrt{\eta \sin I}}$.
\end{itemize}

For $\eta \gg 1$, the solutions obviously just come from $\cos I \pm \sin I \cot
\theta = 0$, which are just $\sin (I \pm \theta) = 0$

@@

We notate $\mu = \cos\theta$; note that CS4 is $\mu_4 \approx \frac{\eta \cos
I}{1 - \eta \sin I} \approx \eta \cos I$. Setting the Hamiltonian equal to its
value at CS4 gives
\begin{align*}
    H_4 &\equiv H\p*{\mu_4, \phi_4}
        \approx -\frac{\mu_4^2}{2} + \eta \mu_4 \cos I - \eta \sin I,\\
        &= +\eta^2\cos^2 I - \eta \sin I,\\
    H(\mu_{sep}, \phi_{sep})
        &= H_4 = -\eta \sin I \cos \phi_{sep} - \frac{\mu_{sep}^2}{2} + \eta
            \mu_{sep} \cos I + \mathcal{O}(\eta^3),\\
    0 &\approx \frac{\mu_{sep}^2}{2} - \eta \mu_{sep} \cos I
        - \eta \sin I\p*{1 - \cos \phi_{sep}} + \eta^2\cos^2I,\\
    \mu_{sep}\p*{\phi} &\approx \sqrt{2\eta \sin I\p*{1 - \cos \phi}}
        + \mathcal{O}(\eta).
\end{align*}

@@

\subsection{Try 1: Qualitative}

We zoom in on Cassini State $4$, which has $\theta_4 =
-\frac{\pi}{2} + \frac{\eta \cos I}{1 - \eta \sin I}, \mu_4 = \frac{\eta \cos
I}{1 - \eta \sin I}, \phi_4 = 0$. Then, using
equations of motion
\begin{align}
    \pd{\phi}{t} &= \mu - \eta\p*{\cos I + \sin I \frac{\mu}{\sqrt{1 - \mu^2}}
        \cos \phi} ,\\
    \pd{\mu}{t} &= -\eta \sin I \sin \phi + \s*{\epsilon \p*{1 - \mu^2}},
\end{align}
we can perturbatively require $\pd{\theta}{t} = 0$ for $\epsilon \neq 0$. This
corresponds to $\eta \sin I \sin \p*{\phi_4 + \delta \phi} \approx \epsilon$, or
$\delta \phi_4 = +\frac{\epsilon}{\eta \sin I}$. This is in agreement with
Dong's result. Note that $\delta \theta_2 = -\frac{\epsilon}{\eta \sin I}$,
which I saw in my simulations.

@@

\subsection{Try 2: Melnikov Distance}

We notice that the separatrix is a heteroclinic orbit, or a saddle connection,
in the dissipation free problem. Introducing dissipation breaks the saddle
connection by a distance that can be estimated with the Melnikov distance. This
is G\&H Equation 4.5.11 or something:
\begin{align}
    d(t_0) &= \frac{\epsilon M(t_0)}{\abs*{f(q^0(0))}} + \mathcal{O}(\epsilon^2)
        ,\\
    M(t_0) &= \int\limits_{-\infty}^\infty
        \s*{f \times g}_{hetero}\;\mathrm{d}t.
\end{align}

This is not a hard formula to understand; along the separatrix, motion is
dominated by $f$, but the perpendicular component adds up to contribute to a
total ``perpendicular distance away from the original separatrix'' necessary to
hit the saddle point, at least intuitively.

We evaluate the Melnikov integral $M(t_0)$ on the heteroclinic orbit. Note that
since in our problem our perturbation $g$ is time-independent, so too is the
Melnikov integral $M(t_0) = M$.

Let's apply this to the Cassini state Hamiltonian w/ dissipation. We first write
down our EOM in Melnikov form (we use canonical variables $\mu, \phi$):
\begin{align}
    \rd{\hat{s}}{t}
        &= \underbrace{
            \pd{\mathcal{H}}{\mu}\hat{\phi} - \pd{\mathcal{H}}{\phi}\hat{\mu}}_f
            + \epsilon \underbrace{\p*{1 - \mu^2}\hat{\mu}}_g.
\end{align}
Then $f \times g = f_\phi g_\mu = \pd{\mathcal{H}}{\mu}\p*{1 - \mu^2}$. We then
want to integrate this along the heteroclinic orbit. We can make change of
variables
\begin{equation}
    M = \int\limits_0^{2\pi}\pd{\mathcal{H}}{\mu}
        \p*{1 - \mu^2}\p*{\pd{\phi}{t}}^{-1}\;\mathrm{d}\phi.
\end{equation}
But thankfully, $\pd{\mathcal{H}}{\mu} = \pd{\phi}{t}$ in the absence of
dissipation, and so $M = 2\pi\p*{1 - \mu^2} \approx 2\pi\p*{1 - 2\eta \sin I}$.
Thus, the Melnikov distance at point $q^0$, a point on the heteroclinic
orbit of the unperturbed Hamiltonian, is just
\begin{equation}
    d(q^0) = \frac{2\pi \epsilon\p*{1 - 2\eta \sin I}}{\abs*{f(q^0)}}.
\end{equation}
Note that the maximum value $\abs*{f(q^0)}$, which occurs at $\phi = \pi$, is
just $f \approx \sqrt{4\eta \sin I}$.

It proves to be a bit difficult to make quantitative predictions though, since
the phase diagram is very smushed where $f$ is large, and $d$ is rather
inaccurate where $f$ is small. Let's think about a Poincar\'e map instead.

\subsection{Try 3: Poincar\'e Section}

Let's consider the Poincar\'e section every time $\phi = \phi_4$ as the
trajectory subject to tidal dissipation is moving $\theta < \theta_4 \to
\theta_4$. To provide an estimate of $\Delta \theta(\theta) = \theta_{n - 1} -
\theta_n$, this is just $\epsilon T$ where $T$ is the time elapsed between
$\theta_n, \theta_{n + 1}$, the period of the orbit. $T$ is dominated by when
$\pd{\phi}{t} \ll 1$ though, or where the orbit is close to the saddle point.

Consider when $\mu = \mu_4 - \delta \mu$
